Event-Driven Simulation of Foraging Dynamics

### 2016-07-21 Manuscript/Simulation_description/simulation-new.tex
cleaning up old files

According to these processes and also under the assumption that the densities
of full foragers, hungry foragers, and resources (also denoted by $F$,
$H$, and $R$, respectively) are perfectly mixed, they evolve according to the
rate equations:
\begin{align}
  \label{RE}
\begin{split}
\dot F &= \lambda F + \rho  RH - \sigma (1-R)F\,,\\
\dot H &= \sigma (1-R)F - \rho RH - \mu H\,, \\
\dot R &= \alpha R(1-R) -  R(F+H),\\
\end{split}
\end{align}
where the overdot denotes time derivative.

We now outline an event-driven algorithm that mimic these rate equations.  We
will also generalize to the situation where the full and hungry foragers
undergo diffusion with possibly different diffusion coefficients on a
finite-dimensional lattice.  Suppose that the system at some time consists of
$N_F$ full foragers, $N_H$ hungry foragers, and $N_R$ individual resources.
The total number of particles $N=N_F+N_H+N_R$.

In each elemental event an individual is picked: a full forager with
probability $N_F/N$, a hungry forager with probability $N_H/N$, and a
resource with probability $N_R/N$.  If an $F$ is picked, it reproduces at
rate $\lambda$ and becomes hungry with rate $\sigma(1-R)$.  If an $H$ is
picked, it becomes full with rate $\rho R$ and dies with rate $\mu$.
Finally, if an $R$ is picked, it grows with rate $\alpha(1-R)$ and is eaten
with rate $(F+H)$.  Thus the total rate for each elemental event is
\begin{equation}
\mathcal{R}=F\big[\lambda +\sigma(1-R)\big]+ H\big[\rho R+\mu\big]+
R\big[\alpha(1-R)+(F+H)\big]\,.
\end{equation}

@@

We now determine the change in the expected number of individuals of each
type in a single event that are consistent with the above processes.  For
full foragers, this change is
\begin{subequations}
\begin{equation}
  \Delta N_F =\left[\frac{N_F}{N}\big(\lambda-\sigma(1-R)\big)+\frac{N_H}{N}\rho
      R\right]/\mathcal{R}\,.
\end{equation}
The term proportional to $N_F$ comes from processes in which a full forager
is picked while the term proportional to $N_H$ comes from processes in which
a hungry forager is picked and it converted to a full forager.  Consequently,
the change in the density of full foragers simply is
\begin{equation}
  \Delta F =\left[F\big(\lambda-\sigma(1-R)\big)+H\rho
    R\right]/N\mathcal{R}\,.
\end{equation}
\end{subequations}
Thus if we take the time step for each event to be
$\Delta t = (N\mathcal{R})^{-1}$, the above reduces to the rate equation
\eqref{RE} for $F$.  Thus in each microscopic event of the model, the time
should be advanced by $\Delta t = (N\mathcal{R})^{-1}$.

In a similar fashion, the change in the expected number of hungry foragers in
a single event is given by
\begin{subequations}
\begin{equation}
  \Delta N_H =\left[-\frac{N_H}{N}(\rho R+\mu)+\frac{N_F}{N}\sigma
    (1-R)\right]/\mathcal{R}\,,
\end{equation}
so that the change in the density of hungry foragers simply is
\begin{equation}
  \Delta H =\left[-H(\rho R+\mu)+F\sigma (1-R)\right]/N\mathcal{R}\,.
\end{equation}
\end{subequations}
Finally, the change in the expected number of individual resources in a
single event is given by
\begin{subequations}
\begin{equation}
  \Delta N_R =\left[\frac{N_R}{N}\big(\alpha(1-R)\big)-\frac{(N_F+N_H)}{N}\right]/\mathcal{R}\,,
\end{equation}
so that the change in the density of resources is
\begin{equation}
  \Delta R =\left[R\big(\alpha(1-R)\big)-(F+H)\right]/N\mathcal{R}\,.
\end{equation}
\end{subequations}
The equations for $\Delta F$, $\Delta H$, and $\Delta R$ then reproduce the
original rate equations \eqref{RE} when the time step for an elemental event
is taken to be $\Delta t = (N\mathcal{R})^{-1}$.

@@

We now determine the total rate for an elemental event.  If an $F$ is picked,
it reproduces at rate $\lambda$, becomes hungry with rate $\sigma(1-R)$, and
moves with rate $D_F$.  If an $H$ is picked, it becomes full with rate
$\rho R$, dies with rate $\mu$, and moves with rate $D_H$.  As before, if an
$R$ is picked, it grows with rate $\alpha(1-R)$ and is eaten with rate
$(F+H)$.  Thus the total rate for each elemental event is
\begin{equation}
\mathcal{R}_d=F\big[\lambda +\sigma(1-R)+D_F\big]+ H\big[\rho R+\mu+D_H\big]+
R\big[\alpha(1-r)+(F+H)\big]\,.
\end{equation}
Here the subscript $d$ indicates that this rate pertains to the foraging
model with diffusing foragers on a $d$-dimensional lattice.

@@

Now let's turn to the mechanics of the various update events.  In the
mean-field limit, when a full forager reproduces, its offspring should be
placed anywhere.  For a finite-dimensional lattice, it seems reasonable to
place the offspring at the same site as the parent.  Diffusion of a full
forager on a lattice can be modeled by allowing it to hop to a nearest
neighbor; similarly for the motion of hungry foragers.

>>>>

According to these processes and also under the assumption that the densities
of full foragers, hungry foragers, and resources (also denoted by $F$,
$H$, and $R$, respectively) are perfectly mixed, they evolve according to the
rate equations:
\begin{align}
  \label{RE}
\begin{split}
\dot F &= \lambda F + \rho  RH - \sigma (1-R)F\,,\\
\dot H &= \sigma (1-R)F - \rho RH - \mu H\,, \\
\dot R &= \alpha R(1-R) -  R(F+H),\\
\end{split}
\end{align}
where the overdot denotes time derivative.  An important feature is the
assumption that the carrying capacity is set equal to 1.  This means that
when the system is completely occupied by resources there can be no further
generation of the resource.  The approach described below can be generalized
to carrying capacity less than 1, but one can take the carrying capacity
equal to 1 without loss of generality.

We now outline an event-driven algorithm that mimic these rate equations.
This algorithm is based on picking \emph{sites} at random, rather than
\emph{aprticles} at random.  However, we show how to recast this site-based
simulation into a particle-based simulation with an concomitant increase in
efficiency.  We will also generalize to the situation where the full and
hungry foragers undergo diffusion with possibly different diffusion
coefficients on a finite-dimensional lattice.

Suppose that the system at some time consists of $N_F$ full foragers, $N_H$
hungry foragers, and $N_R$ individual resources.  The total number of
particles $N=N_F+N_H+N_R$.  In each elemental event a lattice site is picked:
on average, it is occupied by a full forager with probability $N_F/V$, a
hungry forager with probability $N_H/V$, a resource with probability $N_R/V$,
or, on average, the site may be empty with probability $1-\rho$, where
$\rho=(N_F+N_H+N_R)/V$ is the total density of particles, on average and $V$
is the total number of lattice sites..  I use the term ``on average'' because
it is possible to have rare realizations where all particles are concentrated
on just a few sites.  However, for large systems, the actually number of
particles at a given site will not signficantly deviate from the average
number.

If a site that is occupied by an $F$ is picked, it reproduces at rate
$\lambda$ and becomes hungry with rate $\sigma(1-R)$.  If a site that is
occupied by an $H$ is picked, it becomes full with rate $\rho R$ and dies
with rate $\mu$.  Finally, if a site that is occupied by an $R$ is picked, it
grows with rate $\alpha(1-R)$ and is eaten with rate $(F+H)$.  Thus the total
rate for each elemental event is
\begin{equation}
\mathcal{R}=F\big[\lambda +\sigma(1-R)\big]+ H\big[\rho R+\mu\big]+
R\big[\alpha(1-R)+(F+H)\big]+(1-\rho)\,.
\end{equation}

@@

We now determine the change in the expected number of individuals of each
type in a single event that are consistent with the above processes.  For
full foragers, this change is
\begin{subequations}
\begin{equation}
  \Delta N_F =\left[\frac{N_F}{V}\big(\lambda-\sigma(1-R)\big)+\frac{N_H}{V}\rho
      R\right]/\mathcal{R}\,.
\end{equation}
The term proportional to $N_F$ comes from processes in which a full forager
is picked while the term proportional to $N_H$ comes from processes in which
a hungry forager is picked and it converted to a full forager.  Consequently,
the change in the density of full foragers simply is
\begin{equation}
  \Delta F =\frac{\Delta N_F}{V} =\left[F\big(\lambda-\sigma(1-R)\big)+H\rho
    R\right]/V\mathcal{R}\,.
\end{equation}
\end{subequations}
Thus if we take the time step for each event to be
$\Delta t = (V\mathcal{R})^{-1}$, the above reduces to the rate equation
\eqref{RE} for $F$.  Thus in each microscopic event of the model, the time
should be advanced by $\Delta t = (V\mathcal{R})^{-1}$.

In a similar fashion, the change in the expected number of hungry foragers in
a single event is given by
\begin{subequations}
\begin{equation}
  \Delta N_H =\left[-\frac{N_H}{V}(\rho R+\mu)+\frac{N_F}{V}\sigma
    (1-R)\right]/\mathcal{R}\,,
\end{equation}
so that the change in the density of hungry foragers simply is
\begin{equation}
  \Delta H =\left[-H(\rho R+\mu)+F\sigma (1-R)\right]/V\mathcal{R}\,.
\end{equation}
\end{subequations}
Finally, the change in the expected number of individual resources in a
single event is given by
\begin{subequations}
\begin{equation}
  \Delta N_R =\left[\frac{N_R}{V}\big(\alpha(1-R)\big)-\frac{(N_F+N_H)}{V}\right]/\mathcal{R}\,,
\end{equation}
so that the change in the density of resources is
\begin{equation}
  \Delta R =\left[R\big(\alpha(1-R)\big)-(F+H)\right]/V\mathcal{R}\,.
\end{equation}
\end{subequations}
The equations for $\Delta F$, $\Delta H$, and $\Delta R$ would then reproduce
the original rate equations \eqref{RE} when the time step for an elemental
event is taken to be $\Delta t = (V\mathcal{R})^{-1}$.

However, this approach is inefficient if many of the lattice sites are empty
because an empty site is likely to be picked and then nothing happens.
Instead, we would rather pick \emph{actual particles} rather than
\emph{lattice sites}.  If the particle density is $\rho=N/V$, then in picking
a \emph{particle} at random always leads to a non-trivial update, while
picking a \emph{site} at random leads to a non-trivial update with
probability $\rho$.  Thus by picking \emph{particles} at random---an $F$ with
probability $N_F/N$, an $H$ with probability $N_H/N$, or an $R$ with
probability $N_R/N$---compared to picking \emph{sites} at random, leads to an
overall speedup of the simulation by a factor of $1/\rho$.  Therefore, for
the particle-based upate, time increment changes to
$\Delta t = (1/\rho)\times (V\mathcal{R})^{-1}= (N\mathcal{R})^{-1}$.

To summarize, pick a particle of type $i$ with $N_i/N$, and update the
densities $N_i/V$ according (3b), (4b), and (5b), and then increment the time
by $(N\mathcal{R})^{-1}$ after each event.  It is important to appreciate
that this appoach implicitly assumes that the number of particles on each
site is small.  This assumption is reasonable if we are describing a harsh
environment where a non-negligible fraction of the foragers are either going
hungry or starving.

@@

We now determine the total rate for an elemental event.  If an $F$ is picked,
it reproduces at rate $\lambda$, becomes hungry with rate $\sigma(1-R)$, and
moves with rate $D_F$.  If an $H$ is picked, it becomes full with rate
$\rho R$, dies with rate $\mu$, and moves with rate $D_H$.  As before, if an
$R$ is picked, it grows with rate $\alpha(1-R)$ and is eaten with rate
$(F+H)$.  Thus the total rate for each elemental event is
\begin{equation}
\mathcal{R}_d=F\big[\lambda +\sigma(1-R)+D_F\big]+ H\big[\rho R+\mu+D_H\big]+
R\big[\alpha(1-r)+(F+H)\big]+(1-\rho)\,.
\end{equation}
Here the subscript $d$ indicates that this rate pertains to the foraging
model with diffusing foragers on a $d$-dimensional lattice.

@@

Now let's turn to the mechanics of the various update events.  In the
mean-field limit, when a full forager reproduces, its offspring should be
placed anywhere.  For a finite-dimensional lattice, it seems reasonable to
place the offspring at the same site as the parent.  However, because the
implicity assumption of low density, perhaps it would be better to put the
newly born forager on the closest lattice site that is not occupied by
another forager.  Diffusion of a full forager on a lattice can be modeled by
allowing it to hop to a nearest neighbor; similarly for the motion of hungry
foragers.
